preciseTAD: A machine learning framework for precise 3D domain boundary prediction at base-level resolution

### 2021-01-21 BMC_Genome_Biology_submission/grubert_manuscript_figures_tables/manuscript/manuscript.Rmd
revised commenting scheme for new sentences/paragraphs

We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We used Arrowhead-called TAD boundaries [@durand2016juicer] and published Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K562 cell lines at 5-100 kb resolution (Arrowhead) and 10 kb resolution (Peakachu) were used (Additional file 1: Arrowhead Script, Additional file 2: Table S1<!--Supplementary Table 1-->). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.

<!--%%%%
We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. For TAD prediction, we derived cell line specific boundaries at 5-100 kb resolutions using *Arrowhead* [@durand2016juicer] (Additional file 1: Arrowhead Script, Additional file 2: Table S1). For chromatin loop prediction, we downloaded loops derived by *Peakachu* [@salameh2019supervised] at 10 kb (Additional file 2: Table S1). Our models were trained on cell line-specific functional genomic annotation data (ChIP-seq peak regions) publicly available through ENCODE [@encode2004encode; @davis2018encyclopedia] (Additional file 4: Table S3). We used ChIA-PET derived loop boundaries [@grubert2020landscape] (Additional file 2: Table S1), referred hereafter as *Grubert* boundaries, as the gold standard when validating our models and extracting performance metrics (See Methods). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.-->

@@

|          Tool         |     Resolution/Bin Size    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:--------------------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb               |                         8052                       |                             15468                           |                535363               |           0.03         |
|     Arrowhead         |       10   kb              |                         7676                       |                             14253                           |                267682               |           0.05         |
|     Arrowhead         |       25   kb              |                         4670                       |                             8363                            |                107073               |           0.08         |
|     Arrowhead         |       50   kb              |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     Arrowhead         |      100   kb              |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        Peakachu       |       10   kb              |                        16185                       |                          21421                              |                267682               |        0.14            |
|        Grubert        |       5    kb              |                        16232                       |                          18455                              |                535363               |        0.07            |
Table: Domain boundary data and class imbalance summaries across resolutions for Arrowhead, Peakachu<!--%%%, and Grubert--> on GM12878.

@@

<!--%%%Our training/testing framework was used to determine the optimal set of data-level characteristics for TAD/loop boundary region prediction including resolution (bin size), feature engineering, and resampling (Figure 2).--> In total, we considered combinations of data from two cell lines *L = {GM12878, K562}*, five resolutions *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor spaces *P = {Signal, OC, OP, Distance}*, and three re-sampling techniques *S = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed at 500 and at 0.1% of the rows in the training data, respectively. <!--%%%Models were validated on the testing data using cell line-specific Grubert-defined boundaries as $\textbf{Y}_{test}$ for gold standard comparison.--> Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model in an unbiased manner.

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from the GM12878 cell line resulted in an average AUC=0.792 <!--%%% 0.840--> (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to 0.795 <!--%%% 0.899-->. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from the GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=0.881 <!--%%% 0.923--> and 0.874 <!--%%% 0.876-->, respectively). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file 7: Figure S3D). The average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

Next, we evaluated the biological significance of *preciseTAD* boundary points. The signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (Figure 5C, 5D; Additional file 10: Figure S5A, S5B <!--%%%Figure 6A, 6B; Additional file 10: Figure S5A, S5B -->). <!--%%%Additionally, we found that when trained on cohesin-mediated chromatin loops, *preciseTAD*-predicted boundaries showed similar levels of enrichment to the set of ChIA-PET derived loop boundaries (Figure 6C; Additional file 10: Figure S5C).--> Surprised by the poor signal distribution around Arrowhead and Peakachu boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, <!--%%%Grubert,--> and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead-, Peakachu-, and SpectralTAD-called boundaries,in contrast to the well-performing Grubert-defined boundaries (Additional file 10: Figure S5C, S5D <!--%%%Additional File 11: Figure S6-->). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone<!--%%%, while showing similar levels of enrichment when compared to Grubert--> (Additional file 11-12: Figure S6-S7 <!--%%%Additional File 12-14: Figure S7-S8-->). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, Figure 6A, Additional file 13: Figure S8). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.<!--%%%While *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS compared to either Arrowhead and Peakachu boundaries alone, Grubert boundaries were found to be the most spatially proximal (Wilcoxon p-value < 0.001; Figure 7A; Additional file 15: Figure S10). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation when compared to a novel TAD-caller or loop prediction algorithm.-->

@@

Previous studies suggest that TAD boundaries are conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu. Only 26% and 49% of boundaries were conserved between cell lines for Arrowhead and Peakachu boundaries (J=0.186 and J=0.388), respectively (Figure 6B, 6D). However, 45%/56% of *preciseTAD*-predicted domain boundaries were conserved between GM12878 and K562 cell lines when using models trained on Arrowhead/Peakachu data (J=0.383 and J=0.444, respectively, Figure 6C, 6E). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.

<!--%%%Previous studies suggest that TAD boundaries are conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, Peakachu, and Grubert. Only 26% and 49% of boundaries were conserved between cell lines for Arrowhead and Peakachu boundaries (J=0.186 and J=0.388),respectively, while Grubert boundaries exhibited the largest level of conservation among TAD/loop calling tools at 52% (J=0.464) (Figure 7C-7D). Comparatively, *preciseTAD*-predicted boundaries were more conserved between GM12878 and K562 cell lines regardless of which ground truth boundary was used (45%; J=0.383, 56%; J=0.444, and 59%; J=0.495 respectively) (Figure 7E-7F). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.-->

@@

The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model (Additional file 14: Figure S9). This is expected as loop boundaries detected by Peakachu are more abundant, while comparatively wide TAD boundaries detected by Arrowhead likely represent the higher level of the 3D chromatin organization. Together with the results in Figure 4C, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.

<!--%%%
The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by either the Peakachu- or Grubert-trained model (Additional file 16: Figure S11). This is expected as loop boundaries detected by Peakachu and Grubert are more abundant, while comparatively wide TAD boundaries detected by Arrowhead likely represent the higher level of the 3D chromatin organization. Together with the results in Figure 4C, the high overlap between TAD and loop boundaries predicted by Arrowhead- and Peakachu-/Grubert-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.-->

Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether models trained using Arrowhead/Peakachu<!--%%%/Grubert--> ground truth data in one cell line could be leveraged to predict boundaries using annotation data from another cell line. We evaluated two scenarios: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs. training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs. training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead-trained models, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries overlapped in both cross-cell-line prediction scenarios (Additional file 15: Figure S10A, S10B <!--Additional file 17: Figure S12-->). When using Peakachu-trained models, we observed 85% (J=0.705) and 88% (J=0.759) overlap (Additional file 15: Figure S10C, S10D <!--Additional file 18: Figure S13-->). <!--%%%Lastly, when using Grubert-trained models, we observed 86% (J=0.698) and 80% (J=0.0.660) overlap (Additional file 19: Figure S14).--> Furthermore, boundaries predicted on unseen annotation data exhibited a similar level of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line (Additional files 17-19: Figure S12-S14). These results indicate that *preciseTAD* pre-trained models can be successfully used to accurately predict both TAD and loop boundaries for cell lines lacking Hi-C data but for which genome annotation data is available.

@@

We present *preciseTAD*, a machine learning approach for precise prediction of TAD and chromatin loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution domain boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against two boundary calling tools <!--%%%multiple boundary calling tools including Arrowhead [@durand2016juicebox], an established TAD-caller, Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops, and Grubert a method for mapping cohesin-mediated loops via chromatin interaction analysis by paired-end tag sequencing (ChIA-PET) [@grubert2020landscape].-->, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between boundary regions and genomic annotations coupled with random under-sampling results in the best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package, provides the pre-trained models, and allows for precise prediction of boundaries for any cell line for which the genomic annotations are available.

<!--%%%%
We present *preciseTAD*, a machine learning approach for precise prediction of TAD and chromatin loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution domain boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against multiple boundary calling tools including Arrowhead [@durand2016juicebox], an established TAD-caller, Peakachu [@salameh2019supervised], a recently published algorithm for predicting chromatin loops, and Grubert [@grubert2020landscape], a method for mapping cohesin-mediated loops via chromatin interaction analysis by paired-end tag sequencing (ChIA-PET). We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between boundary regions and genomic annotations coupled with random under-sampling results in the best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package (version 1.1.1; Bioconductor version 3.12) and provides the pre-trained models, via an ExperimentHub R package *preciseTADhub* (version 0.99.8; Bioconductor version 3.13), which allows for precise prediction of boundaries for any cell line for which the genomic annotations are available.
-->

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (Additional file 16: Table S5 <!--%%%Additional file 20: Table S5-->). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't emphasize one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be susceptible to different values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us from investigating its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons, we opted to report balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data resampling.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 3 different thresholds ($t=(0.975,0.99,1.0)$) and 6 different $\epsilon$ values ($\epsilon=(1000,5000,10000,15000,20000,25000)$)). We found that the normalized overlaps - calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries - between top TFBS sites and flanked *preciseTAD* boundaries converged for combinations of $t=1.0$ and $\epsilon=10000$ (Additional file 17: Figure S11 <!--%%%Additional file 21: Figure S15-->).

@@

<!--%%%
## Predicting boundaries across cell lines

>>>>

<!--
We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We used Arrowhead-called TAD boundaries [@durand2016juicer] and published Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K562 cell lines at 5-100 kb resolution (Arrowhead) and 10 kb resolution (Peakachu) were used (Additional file 1: Arrowhead Script, Additional file 2: Table S1). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.
-->

<!--%%%% revised paragraph-->
We developed a machine learning (ML) framework for determining the optimal set of data level characteristics to predict boundary regions of Topologically Associating Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. For TAD prediction, we derived cell line specific boundaries at 5-100 kb resolutions using *Arrowhead* [@durand2016juicer] (Additional file 1: Arrowhead Script, Additional file 2: Table S1). For chromatin loop prediction, we downloaded loops derived by *Peakachu* [@salameh2019supervised] at 10 kb (Additional file 2: Table S1). Our models were trained on cell line-specific functional genomic annotation data (ChIP-seq peak regions) publicly available through ENCODE [@encode2004encode; @davis2018encyclopedia] (Additional file 4: Table S3). We used ChIA-PET derived loop boundaries [@grubert2020landscape] (Additional file 2: Table S1), referred hereafter as *Grubert* boundaries, as the gold standard when validating our models and extracting performance metrics (See Methods). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base-level resolution from cell line-specific genome annotation data.
<!--%%%-->

@@

|          Tool         |     Resolution/Bin Size    |     Total number of called TADs/chromatin loops    |     Total number of unique TAD/chromatin loop boundaries    |     Total number of genomic bins    |     Class imbalance    |
|:---------------------:|:--------------------------:|:--------------------------------------------------:|:-----------------------------------------------------------:|:-----------------------------------:|:----------------------:|
|        Arrowhead      |       5   kb               |                         8052                       |                             15468                           |                535363               |           0.03         |
|     Arrowhead         |       10   kb              |                         7676                       |                             14253                           |                267682               |           0.05         |
|     Arrowhead         |       25   kb              |                         4670                       |                             8363                            |                107073               |           0.08         |
|     Arrowhead         |       50   kb              |                         2349                       |                             4224                            |                 53537               |           0.08         |
|     Arrowhead         |      100   kb              |                         1031                       |                             1883                            |                 26768               |           0.07         |
|        Peakachu       |       10   kb              |                        16185                       |                          21421                              |                267682               |        0.14            |
|        Grubert        |       5    kb              |                        16232                       |                          18455                              |                535363               |        0.07            |
Table: Domain boundary data and class imbalance summaries across resolutions for Arrowhead, Peakachu<!--%%%-->, and Grubert<!--%%%--> on GM12878.

@@

<!--%%%-->Our training/testing framework was used to determine the optimal set of data-level characteristics for TAD/loop boundary region prediction including resolution (bin size), feature engineering, and resampling (Figure 2).<!--%%%--> In total, we considered combinations of data from two cell lines *L = {GM12878, K562}*, five resolutions *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor spaces *P = {Signal, OC, OP, Distance}*, and three re-sampling techniques *S = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed at 500 and at 0.1% of the rows in the training data, respectively. <!--%%%-->Models were validated on the testing data using cell line-specific Grubert-defined boundaries as $\textbf{Y}_{test}$ for gold standard comparison.<!--%%%--> Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model in an unbiased manner.

@@

Having demonstrated the optimal performance in our framework's ability to predict boundary regions, we next attempted to predict boundary regions in one cell line using the model pre-trained on data from another cell line (See Methods). We found that training and testing using Arrowhead ground truth TAD boundaries and genomic annotation data from the GM12878 cell line resulted in an average AUC=<!--0.792--> <!--%%%--> 0.840<!--%%%--> (Figure 4D). Interestingly, when training on the K562 cell line and testing on GM12878, the average AUC increased slightly to <!--0.795--> <!--%%%--> 0.899<!--%%%-->. Likewise, the average performance of models trained using Peakachu boundaries and genomic annotation data from the GM12878 cell line was comparable to models trained on K562-specific Peakachu boundaries and genomic annotations (Avg. AUC=<!--0.881--> <!--%%%--> 0.923<!--%%%--> and <!--0.874--> <!--%%%--> 0.876<!--%%%-->, respectively). These results were consistent when comparing training/testing strategies on K562 with training on GM12878 and testing on K562 data (Additional file 7: Figure S3D). The average ROC curves were found to be within 1 standard deviation of each other, suggesting that a model trained on data from one cell line performs well when using the data from another cell line, allowing for the opportunity to predict boundaries for cell lines that do not currently have Hi-C data available.

@@

Next, we evaluated the biological significance of *preciseTAD* boundary points. The signal of four known molecular drivers of 3D chromatin (CTCF, RAD21, SMC3, and ZNF143) colocalized more frequently around *preciseTAD*-predicted boundary points, as compared to Arrowhead-called TAD and Peakachu loop boundaries, respectively (<!--Figure 5C, 5D; Additional file 10: Figure S5A, S5B--> <!--%%%-->Figure 6A, 6B; Additional file 10: Figure S5A, S5B <!--%%%-->). <!--%%%-->Additionally, we found that when trained on cohesin-mediated chromatin loops, *preciseTAD*-predicted boundaries showed similar levels of enrichment to the set of ChIA-PET derived loop boundaries (Figure 6C; Additional file 10: Figure S5C).<!--%%%--> Surprised by the poor signal distribution around Arrowhead and Peakachu boundaries, we compared signals centered on boundaries called by Arrowhead, Peakachu, <!--%%%-->Grubert,<!--%%%--> and our SpectralTAD [@Cresswell:2020ab] TAD caller. We confirmed the poor signal distribution around Arrowhead-, Peakachu-, and SpectralTAD-called boundaries,in contrast to the well-performing Grubert-defined boundaries (<!--Additional file 10: Figure S5C, S5D--> <!--%%%-->Additional File 11: Figure S6<!--%%%-->). Signal enrichment heatmaps confirmed that *preciseTAD*-predicted boundaries were more enriched for the same genomic annotations than either Arrowhead or Peakachu boundaries alone<!--%%%-->, while showing similar levels of enrichment when compared to Grubert<!--%%%--> (<!--Additional file 11-12: Figure S6-S7--> <!--%%%-->Additional File 12-14: Figure S7-S9<!--%%%-->). *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS (Wilcoxon p-value < 0.001 versus Arrowhead and Peakachu boundaries, respectively, <!--Figure 6A, Additional file 13: Figure S8-->  <!--%%%-->Figure 7A; Additional file 15: Figure S10<!--%%%-->). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation.<!--%%%-->While *preciseTAD* boundaries were statistically significantly closer to the top-ranked TFBS compared to either Arrowhead and Peakachu boundaries alone, Grubert boundaries were found to be the most spatially proximal (Wilcoxon p-value < 0.001; Figure 7A; Additional file 15: Figure S10). These results indicate that *preciseTAD*-predicted boundaries better reflected the known biology of boundary formation when compared to a novel TAD-caller or loop prediction algorithm.<!--%%%-->

@@

<!--
Previous studies suggest that TAD boundaries are conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, and Peakachu. Only 26% and 49% of boundaries were conserved between cell lines for Arrowhead and Peakachu boundaries (J=0.186 and J=0.388), respectively (Figure 6B, 6D). However, 45%/56% of *preciseTAD*-predicted domain boundaries were conserved between GM12878 and K562 cell lines when using models trained on Arrowhead/Peakachu data (J=0.383 and J=0.444, respectively, Figure 6C, 6E). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.
-->

<!--%%%revised paragraph-->
Previous studies suggest that TAD boundaries are conserved across cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @sexton2012three]. To assess the level of cross-cell-line conservation, we evaluated the overlap between cell line-specific boundaries detected by *preciseTAD*, Arrowhead, Peakachu, and Grubert. Only 26% and 49% of boundaries were conserved between cell lines for Arrowhead and Peakachu boundaries (J=0.186 and J=0.388),respectively, while Grubert boundaries exhibited the largest level of conservation among TAD/loop calling tools at 52% (J=0.464) (Figure 7C-7D). Comparatively, *preciseTAD*-predicted boundaries were more conserved between GM12878 and K562 cell lines regardless of which ground truth boundary was used (45%; J=0.383, 56%; J=0.444, and 59%; J=0.495 respectively) (Figure 7E-7F). The better conservation of *preciseTAD*-predicted boundaries further supports the notion of their higher biological relevance.
<!--%%%-->

@@

<!--
The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by the Peakachu-trained model (Additional file 14: Figure S9). This is expected as loop boundaries detected by Peakachu are more abundant, while comparatively wide TAD boundaries detected by Arrowhead likely represent the higher level of the 3D chromatin organization. Together with the results in Figure 4C, the high overlap between boundaries predicted by Arrowhead- and Peakachu-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms. 
-->

<!--%%%revised paragraph-->
The majority of boundaries predicted by the Arrowhead-trained *preciseTAD* model represented a subset of the larger group of boundaries predicted by either the Peakachu- or Grubert-trained model (Additional file 16: Figure S11). This is expected as loop boundaries detected by Peakachu and Grubert are more abundant, while comparatively wide TAD boundaries detected by Arrowhead likely represent the higher level of the 3D chromatin organization. Together with the results in Figure 4C, the high overlap between TAD and loop boundaries predicted by Arrowhead- and Peakachu-/Grubert-trained models suggests that TAD and loop boundaries may be driven by similar molecular mechanisms.
<!--%%%-->

Guided by the high predictive performance when training and testing on different cell lines (Figure 4D), we opted to evaluate whether models trained using Arrowhead/Peakachu<!--%%%-->/Grubert<!--%%%--> ground truth data in one cell line could be leveraged to predict boundaries using annotation data from another cell line. We evaluated two scenarios: 1) training on GM12878 and predicting boundaries on GM12878 (GM on GM) vs. training on K562 and predicting on GM12878 (K on GM), and 2) training on K562 and predicting boundaries on K562 (K on K) vs. training on GM12878 and predicting boundaries on K562 (GM on K). Using Arrowhead-trained models, 76% (J=0.701) and 81% (J=0.751) of predicted boundaries overlapped in both cross-cell-line prediction scenarios (<!--Additional file 15: Figure S10A, S10B--> <!--%%%-->Additional file 17: Figure S12<!--%%%-->). When using Peakachu-trained models, we observed 85% (J=0.705) and 88% (J=0.759) overlap (<!--Additional file 15: Figure S10C, S10D--> <!--%%%-->Additional file 18: Figure S13<!--%%%-->). <!--%%%-->Lastly, when using Grubert-trained models, we observed 86% (J=0.698) and 80% (J=0.0.660) overlap (Additional file 19: Figure S14).<!--%%%--> Furthermore, boundaries predicted on unseen annotation data exhibited a similar level of enrichment for CTCF, RAD21, SMC3, and ZNF143 as did those trained and predicted on the same cell line (Additional files 17-19: Figure S12-S14). These results indicate that *preciseTAD* pre-trained models can be successfully used to accurately predict both TAD and loop boundaries for cell lines lacking Hi-C data but for which genome annotation data is available.

@@

<!--
We present *preciseTAD*, a machine learning approach for precise prediction of TAD and chromatin loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution domain boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against two boundary calling tools, Arrowhead [@durand2016juicebox], an established TAD-caller, and Peakachu [@salameh2019supervised] a recently published algorithm for predicting chromatin loops. We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between boundary regions and genomic annotations coupled with random under-sampling results in the best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package, provides the pre-trained models, and allows for precise prediction of boundaries for any cell line for which the genomic annotations are available.
-->

<!--%%%%revised paragraph-->
We present *preciseTAD*, a machine learning approach for precise prediction of TAD and chromatin loop boundaries from functional genomic annotations. *preciseTAD* leverages a random forest (RF) classification model built on low-resolution domain boundaries obtained from domain calling tools, and high-resolution genomic annotations as the predictor space. *preciseTAD* predicts the probability of each base being a boundary, and identifies the precise location of boundary regions and the most likely boundary points. *preciseTAD* was benchmarked against multiple boundary calling tools including Arrowhead [@durand2016juicebox], an established TAD-caller, Peakachu [@salameh2019supervised], a recently published algorithm for predicting chromatin loops, and Grubert [@grubert2020landscape], a method for mapping cohesin-mediated loops via chromatin interaction analysis by paired-end tag sequencing (ChIA-PET). We first optimized our RF model by systematically comparing different combinations of genome binning (resolution), feature engineering procedures, and resampling techniques. Our results demonstrate that genomic distance between boundary regions and genomic annotations coupled with random under-sampling results in the best model performance. We show that binding of four transcription factors (SMC3, RAD21, CTCF, ZNF143) is sufficient for accurate boundary predictions. *preciseTAD* is implemented as an R package (version 1.1.1; Bioconductor version 3.12) and provides the pre-trained models, via an ExperimentHub R package *preciseTADhub* (version 0.99.8; Bioconductor version 3.13), which allows for precise prediction of boundaries for any cell line for which the genomic annotations are available.
<!--%%%%-->

@@

Besides balanced accuracy (BA), we investigated five other performance metrics, including accuracy, area under the receiver operating characteristic curve (AUROC), precision, F1-score, and area under the precision-recall curve (AUPRC) (<!--Additional file 16: Table S5--> <!--%%%-->Additional file 20: Table S5<!--%%%-->). Our aim was to have a balanced metric sensitive to class imbalance such that it would not favor one component of the confusion matrix. The accuracy metric can be artificially inflated by true negatives (TN), the set of genomic bins correctly predicted as not containing a ground truth boundary. AUROC captures how a model generally performs across different thresholds. However, it doesn't emphasize one class over the other, so it does not reflect the minority class well. Precision indicates the rate at which positive predictions are correct and can be artificially deflated by low proportions of true positives (TP), the set of genomic bins correctly predicted as containing a ground truth boundary. While F1-score is a composite metric, it can be susceptible to different values for precision and recall. Lastly, AUPRC is insensitive toward class imbalance, preventing us from investigating its effect, and also omits from its calculation TN values. All of these are important considerations to make when choosing a performance metric. For these reasons, we opted to report balanced accuracy (BA). The BA benefits from incorporating all components of the confusion matrix, while also being sensitive to class imbalance, a necessary characteristic when comparing performances to models built using no data resampling.

@@

*preciseTAD* offers flexibility in controlling both the number of predicted boundaries and the distance between them. The two primary parameters are the probability threshold $t$ and $\epsilon$ (referred to as eps; parameter of DBSCAN). The combination of these two quantities changes the resulting number of predicted boundaries from *preciseTAD*. Lower values of $t$ and $\epsilon$ will result in more clusters of bases, and therefore, more boundaries. As a heuristic, we evaluated the pairwise combination of 3 different thresholds ($t=(0.975,0.99,1.0)$) and 6 different $\epsilon$ values ($\epsilon=(1000,5000,10000,15000,20000,25000)$)). We found that the normalized overlaps - calculated as the total number of ChIP-seq peaks that overlapped within a given flanked boundary, divided by the number of boundaries - between top TFBS sites and flanked *preciseTAD* boundaries converged for combinations of $t=1.0$ and $\epsilon=10000$ (<!--Additional file 17: Figure S11--> <!--%%%-->Additional file 21: Figure S15<!--%%%-->).

@@

<!--%%%new paragraph-->
## Predicting boundaries across cell lines

We extended the *preciseTAD* algorithm by implementing a strategy to predict domain boundaries across cell lines. To do so, we applied *preciseTAD* on models built using cell line-specific annotation data to predict boundaries on base-level annotation data from another cell line. We compared this strategy to applying *preciseTAD* on models and base level annotation data from the same cell line. Results were compared by assessing the overlap between flanked same-cell-line and cross-cell-line predicted boundaries using venn diagrams and Jaccard indices.
<!--%%%-->
